Fitness tracking reveals task-specific associations between memory, mental health, and exercise

### 2021-10-22 paper/main.tex
addressed all coauthor comments, additional round of JRM edits, updated figs

\begin{abstract}
 Physical exercise can benefit both physical and mental well-being.
  Different forms of exercise (i.e., aerobic versus anaerobic; running
  versus walking versus swimming versus yoga; high-intensity interval
  training versus endurance workouts; etc.) impact physical fitness
  in different ways.  For example, running may substantially impact
  leg and heart strength but only moderately impact arm strength. We
  hypothesized that the mental benefits of exercise might be similarly
  differentiated.  We focused specifically on how different forms of
  exercise might related to different aspects of memory and mental
  health.  To test our hypothesis, we collected roughly a century's
  worth of fitness data (in aggregate).  We then asked participants to
  fill out surveys asking them to self-report on different aspects of
  their mental health.  We also asked participants to engage in a
  battery of memory tasks that tested their short and long term
  episodic, semantic, and spatial memory.  We found that participants
  with similar exercise habits and fitness profiles tended to also
  exhibit similar mental health and task performance profiles.
\end{abstract}

\section*{Introduction}
Engaging in physical activity (exercise) can improve our physical
fitness by increasing muscle strength~\citep{RogeEvan93, Lind79,
  CranEtal13, Knut07}, increasing bone density~\citep{ChilEtal12,
  BassRams94, LaynNels99}, increasing cardiovascular
performance~\citep{MaioEtal00, PollEtal00}, increasing lung
capacity~\citep{LazoEtal16}~\citep[although see][]{RomaEtal16},
increasing endurance~\citep{WilmKnut03}, and more.  Exercise can also
improve mental health~\citep{Ragl90, MikkEtal17, TaylEtal85,
  DeslEtal09, Call04, PaluSchw00, BassSuzu17} and cognitive performance~\citep{ChanEtal12b,
  BrisEtal02, EtniEtal06, BassSuzu17}.

@@

In the present study, we hypothesize that (a) different exercise regimens will
have different, quantifiable impacts on cognitive performance and
mental health, and that (b) these impacts will be consistant across
individuals.  To this end, we collected a year of fitness tracking
data from each of 113 participants.  We then asked each participant to
fill out a brief survey in which they self-evaluated several aspects
of their mental health.  Finally, we ran each participant through a
battery of memory tasks, which we used to evaluate their memory
performance along several dimensions.  We examined the data for
potential associations between memory, mental health, and exercise.

@@

    We ran an online experiment using the Amazon Mechanical Turk
    platform.  We collected data about each participant's fitness and
    exercise habits, a variety of self-reported measures concerning their
    mental health, and about their performance on a battery of memory
    tasks.  We mined the dataset for potential associations between
    memory, mental health, and exercise.

\subsection*{Experiment}
\subsubsection*{Participants}
We recruited experimental participants by posting our experiment as a
Human Intelligence Task (HIT) on the Amazon Mechanical Turk platform.
We limited participation to Mechanical Turk Workers who had been
assigned a ``Masters'' designation on the platform, given to workers
who score highly across several metrics on a large number of HITs,
according to a proprietary algorithm managed by Amazon.  We further
limited our participant pool to participants who self-reported that
they were fluent in English and regularly used a Fitbit fitness
tracker device.  A total of 160 workers accepted our HIT in order to
participate in our experiment.  Of these, we excluded all participants
who failed to log into their Fitbit account (giving us access to their
anonymized fitness tracking data), encountered technical issues (e.g.,
by accessing the HIT using an incompatible browser, device, or
operating system), or who ended their participation prematurely,
before completing the full study.  In all, 113 participants
contributed usable data to the study.

@@

\paragraph{Reported medical conditions and medications.}
Participants reported having and/or taking medications pertaining to the following medical conditions: anxiety or
depression (4 participants), recent head injury (2 participants), high
blood pressure (1 participant), bipolar (1 participant),
hypothyroidism (1 participant), and other unspecified conditions or medications (1
participant).  Participants reported their current and typical stress
levels on a Likert scale as very relaxed (-2), a little relaxed (-1),
neutral (0), a little stressed (1), or very stressed (2).  The
``current'' stress level reflected participants' stress at the time
they participated in the experiment.
Their responses
ranged from -2 to 2 (current stress: 25\textsuperscript{th} percentile: -2;
50\textsuperscript{th} percentile: -1; 75\textsuperscript{th}
percentile: 1; typical stress: 25\textsuperscript{th} percentile: 0;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 1).  Participants also reported their current level of
alertness on a Likert scale as very sluggish (-2), a little sluggish
(-1), neutral (0), a little alert (1), or very alert (2).  Their
responses ranged from -2 to 2 (25\textsuperscript{th} percentile: 0;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 2).  Nearly all (111 out of 113) participants reported
that they had normal color vision, and 15 participants reported
uncorrected visual impairments (including dyslexia and uncorrected
near- or far-sightedness).

@@

\subsubsection*{Tasks}
Upon accepting the HIT posted on Mechanical Turk, each worker was
directed to read and fill out a screening and consent form, and to
share access to their anonymized Fitbit data via their Fitbit account.
After consenting to participate in our stidy and successfully sharing their Fitbit
data, participants filled out a survey and then engaged in a series of
memory tasks (Fig.~\ref{fig:tasks}).  All stimuli and code for running
the full Mechanical Turk experiment may be found
\href{https://github.com/ContextLab/brainfit-task}{\underline{here}}.

@@

The word lists participants studied were drawn from the categorized
lists reported by \cite{ZimaEtal18}.  Each participant was assigned
four unique randomly chosen lists (in a randomized order), selected
from a full set of 16 lists.  Each chosen list was then randomly
shuffled before presenting the words to the participants.

Participants also performed a final delayed memory test where they
were given 180~s to type out any words they remembered from
\textit{any} of the 4 lists they had studied.

@@

\paragraph*{Naturalistic recall (Fig.~\ref{fig:tasks}b).}
Participants watched a 2.5~minute video clip entitled ``The Temple of
Knowledge.''  The video comprises an animated story told to StoryCorps
by Ronald Clark, who was interviewed by his daughter, Jamilah Clark.
The narrator (Ronald) discusses growing up living in an apartment over
Washington Heights branch of the New York Public Library, where his
father worked as a custodian during the 1940s.

After watching the video clip, participants were asked to type out
anything they remembered about what happened in the video.  They typed
their responses into a text box, one sentence at a time.  When the
participant pressed the return key or typed any final punctuation mark
(``.'', ``!'', or ``?'') the text currently entered into the box was
``submitted'' and added to their transcript, and the text box was
cleared to prevent further editing of any already-submitted text.
This was intended to prevent participants from retroactively editing
their previous responses.  Participants were given up to 10~minutes to
enter their responses.  After 4~minutes participants were given the
option of ending the response period early, e.g., if they felt they
had finished entering all of the information they remembered.  Each
participant's transcript was constructed from their submitted
responses by combining the sentences into a single document and
removing extraneous whitespace characters.

Following this 4--10 minute free response period, participants were
given a series of 10 multiple choice questions about the conceptual
content of the story.  All participants received the same questions,
in the same order.

Participants also performed a final delayed memory test, where they
carried out the free response recall task a second time, near the end
of the testing session.  This resulted in a second transcript, for
each participant.

\paragraph*{Foreign language flashcards (Fig.~\ref{fig:tasks}c).}
Participants studied a series of 10 English-Gaelic word pairs in a
randomized order.  We selected the Gaelic language both for its
relatively small number of native speakers and for its dissimilarity
to other commonly spoken languages amongst Mechanical Turk Workers.
We verified (via self report) that all of our participants were fluent
in English and that they were neither fluent nor familiar with Gaelic.

@@

The fitness tracking data associated with participants' Fitbit
accounts varied in scope and duration according to which device the
participant owned (Fig.~\devices), how often the participant
wore (and/or synced) their tracking device, and how long they had
owned their device.  For example, while all participants' devices
supported basic activity metrics such as daily step counts, only a
subset of the devices with heart rate monitoring capabilities provided
information about workout intensity, resting heart rate, and other
related measures.

Across all devices, we collected the following information: heart rate
data, sleep tracking data, logged bodyweight measurements, logged
nutrition measurements, Fitbit account and device settings, and
activity metrics.

@@

\paragraph{Nutrition.} If available, we extracted any
nutrition-related information affiliated with participants' Fitbit
accounts within 1 year prior to enrolling in our study. Depending on
their specific account settings and usage behaviors, this included a
log of the specific foods they had eaten (and logged) over the past
year, and the amount of water consumed each day.

@@

\subsubsection*{Comparing recent versus baseline measurements.}
We were interested in separating out potential associations between
\textit{absolute} fitness metrics and \textit{relative} metrics.  To
this end, in addition to assessing potential raw (absolute) fitness
metrics, we also defined a simple measure of recent changes in those
metrics, relative to a baseline:
\[
  \Delta_{R, B} m = \frac{B \sum_{i = 1}^R
  m(i)}{R \sum_{i=R + 1}^{R+B}m(i)},
\]
where $m(i)$ is the value of metric $m$ from $i - 1$ days prior to
testing (e.g., $m(1)$ represents the value of $m$ on the day the
participant accepted the HIT, and $m(10)$ represents the value of $m$
9 days prior to accepting the HIT.  Unless otherwise noted, we set
$R = 7$ and $B = 30$.  In other words, to estimate recent changes in
any metric $m$, we divided the average value of $m$ taken over the
prior week by the average value of $m$ taken over the 30 days before
that.

\subsubsection*{Exploratory correlation analyses}
We used a bootstrap procedure to identify reliable correlations
between different memory-related, fitness-related, and
demographic-related variables.  For each of $N = 10,000$ iterations, we
selected (with replacement) a sample of 113 participants to include.
This yielded, for each iteration, a sampled ``data matrix'' with one
row per sampled participant and one column for each measured variable.
When participants were sampled multiple times in a given iteration, as
was often the case, this matrix contained duplicate rows.  Next, we computed the Pearson's correlation
between each pair of columns.  This yielded, for each pair of columns,
a distribution of $N$ bootstrapped correlation coefficients.  If fewer
than $97.5\%$ of the coefficients for a given pair of columns had the
same sign, we excluded the pair from further analysis and considered
the expected correlation between those columns to be undefined.  If
$\geq 97.5\%$ of the coefficients for a given pair of columns had the
same sign (corresponding to a bootstrap-estimated two-tailed $p$
threshold of 0.05), we computed the expected correlation coefficient as:
\[
  \mathbb{E}_{i, j}\left[ r\right] = \tanh\left(\frac{1}{N} \sum_{n=1}^N
  \tanh^{-1}(\mathrm{corr}\left(m(i)_n, m(j)_n\right))\right),
\]
where $m(x)_n$ represents column $x$ of the bootstrapped data matrix
for iteration $n$, $\tanh$ is the hyperbolic tangent, and $\tanh^{-1}$
is the inverse hyperbolic tangent.  We estimated the corresponding $p$-values
for these correlations as one minus the proportion of bootstrapped
correlations with the same sign, multiplied by two.

\subsubsection*{Reverse correlation analyses}
We sought to characterize potential associations between the history
of participants' fitness-related activities leading up to the time
they participated in a memory task and their performance on the given
task.  For each fitness-related variable, we constructed a timeseries
matrix whose rows corresponded to timepoints (sampled once per day)
leading up to the day the participant accepted the HIT for our study,
and whose columns corresponded to different participants.  These
matrices often contained missing entries, since different
participants' Fitbit devices tracked fitness-related activities
differently.  For example, participants whose Fitbit devices lacked
heart rate sensors would have missing entries for any heart
rate-related variables.  Or, if a given participant neglected to wear
their fitness tracker on a particular day, the column corresponding to
that participant would have missing entries for that day.  To create
stable estimates, we smoothed the timeseries of each fitness measure
using a sliding window of 1 week.  In other words, for each fitness
measure, we replaced the ``observed
value'' for each day with the average values of that measure (when
available) over the 7 day interval ending on the given day.

@@

Given a fitness timeseries matrix, $F$, we computed the weighted
average and weighted standard error of the mean of each row of $F$,
where the weights were given by a particular memory-related variable
(row of $M$).  For example, if $F$ contained participants' daily step
counts, we could use any row of $M$ to compute a weighted average
across any participants who contributed step count data on each day.
Choosing a row of $M$ that corresponded to participants' performance
on the naturalistic recall task would mean that participants who
performed better on the naturalistic recall task would contribute more
to the weighted average timeseries of daily step counts.
Specifically, for each row, $t$, of $F$, we computed the weighted
average (across the $S$ participants) as:
\[
\bar{f}(t) = \sum_{s=1}^S \dot{m}(s) F(t, s),
\]
where $\dot{m}$ denotes the normalized min-max scaling of $m$ (the row
of $M$ corresponding to the chosen memory-related variable):
\[
  \dot{m} = \frac{m}{\sum_{s=1}^S \hat{m}(s)},
\]
where
\[
  \hat{m} = (1 - \epsilon)\frac{m - \min(m)}{\max(m) - \min(m)} + \epsilon.
\]
Here, $\epsilon$ provides a lower bound on the influence of the
lowest-weighted participant's data.  We defined $\epsilon = 0.001$,
ensuring that the lowest-weighted participant had relatively low (but
non-zero) influence.

We computed the weighted standard error of the mean as:
\[
\mathrm{SEM}_m\left(f(t)\right) = \frac{\left| \sum_{s=1}^S \left( F(t, s) -
    \bar{f}(t)\right) \right|}{\sqrt{S}}.
\]

When a given row of $F$ was missing data from one or more
participants, those participants were excluded from the weighted
average for the corresponding timepoint and the weights (across all
remaining participants) were re-normalized to sum to 1.  The above
procedure yielded, for each memory variable, a timeseries of average
(and standard error of the mean) fitness tracking values leading up to
the day of the experiment.

\section*{Results}
Before testing our main hypothesis we examined the behavioral data
from each of four memory tasks: a random word list learning ``free
recall'' task; a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendencies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or exercise.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/behavior_overview_immediate}
\caption{\textbf{Immediate memory tests.}  \textbf{a. Free recall.}
  Left: probability of recalling each word first as a function of its
  presentation position.  Middle: probability of transitioning between
successively recalling the word presented at position $i$, followed by
word presented at position $i + \mathrm{Lag}$.  Right: probability of
recalling each word as a function of its presentation position.  See
Figure~\frDetail~for additional details.
\textbf{b. Naturalistic recall.}  Top: 2D embedding of a 2.5~min video
clip; each dot reflects a narrative event (red denotes early events
and blue denotes later events).  Bottom: 2D embedding of the averaged
transcripts of participants' recountings of the narrative (dots: same
format as top panel).  The arrows denote the average trajectory
directions through the corresponding region of text embedding space,
for any participants whose recountings passed through that region.
Blue arrows denote statistically reliable agreement across
participants ($p < 0.05$, corrected).  See Figure~\natDetail~for
additional details.  \textbf{c. Foreign language
  flashcards.} Each bar denotes the average proportion of correctly
recalled Gaelic-English word pairs from early (first 3), late (last
3), or all (i.e., all 10) study positions.  See
Figure~\vocabDetail~for additional details.\textbf{d. Spatial
  learning.}  Average estimation error in shape locations as a
function of the number of shapes.  See Figure~\spatialDetail~for
additional details.  All panels: error bars and error
ribbons denote bootstrap-estimated 95\% confidence intervals.  Shading
(saturation) denotes results for different subsets of participants
assigned based on their task performance (Figs.~\frDetail,
\natDetail, \vocabDetail, and \spatialDetail~provide information about
which performance metrics and values the shading reflects; in general
more saturated colors denote participants who performed better on the
given task.)  In Panel d, participants are grouped in two ways; in the
left panel, participants are grouped according to the $y$-intercepts of regression lines (estimation error as a
function of the number of shapes); in the right panel, participants
are grouped according to the slopes of the same regression lines.}
\label{fig:immediate_behavior}
\end{figure}

@@

In addition to observing substantial across-participant variability in
memory performance, we also observed substantial variability in
participants' fitness and activity metrics
(Fig.~\ref{fig:fitness_summary}).  We examined recent measurements,
averaged over the week prior to testing
(Fig.~\ref{fig:fitness_summary}a), baselined measurements (average
over the prior week, divided by the average over the preceding 30 day;
Fig.~\ref{fig:fitness_summary}b), along with more gradually varying
measures that tended to remain relatively static over timescales of
weeks to months (Fig.~\ref{fig:fitness_summary}c).
Figure~\fitDists~displays across-participant
distributions for a broad selection of these measures, and Figures~\fitDistgridImmediate,
\fitScatterImmediate, \fitDistgridDelayed, and \fitScatterDelayed~show
different participants' fitness metrics, broken down by their
performance on different memory tasks.

@@

Although the present study cannot reveal causal effects, a large prior
literature provided some insight into potential causal effects by examining the neural and cognitive effects of a variety
of exercise interventions~\citep{ChanEtal15, VidoEtal15, KamiEtal07,
  ImboEtal19, SuwaEtal17, SinhEtal21, Tomp03}.  A limitation of that
prior work is that most of these studies examine how relatively
short-term changes in exercise (e.g., on timescales of hours to days
or, rarely, weeks to months) affect a cognitive performance on single
task or aspect of mental health.  The present study examines
longer-term exercise (over a full year), and relates long-term
exercise history to performance on a variety of tasks and to a variety
of mental health dimensions.

>>>>

\begin{abstract}
  Physical exercise can benefit both physical and mental well-being.
  Different forms of exercise (i.e., aerobic versus anaerobic; running
  versus walking, swimming, or yoga; high-intensity interval training
  versus endurance workouts; etc.) impact physical fitness in
  different ways.  For example, running may substantially impact leg
  and heart strength but only moderately impact arm strength. We
  hypothesized that the mental benefits of exercise might be similarly
  differentiated.  We focused specifically on how different forms of
  exercise might related to different aspects of memory and mental
  health.  To test our hypothesis, we collected (in aggregate) roughly
  a century's worth of fitness data.  We then asked participants to
  fill out surveys asking them to self-report on different aspects of
  their mental health.  We also asked participants to engage in a
  battery of memory tasks that tested their short and long term
  episodic, semantic, and spatial memory performance.  We found that
  participants with similar exercise habits and fitness profiles
  tended to also exhibit similar mental health and task performance
  profiles.  These effects were task-specific in that different
  exercise patterns or fitness characteristics varied with different
  aspects of memory, on different tasks.
\end{abstract}

\section*{Introduction}
Engaging in physical activity (exercise) can improve physical
fitness by increasing muscle strength \citep{RogeEvan93, Lind79,
  CranEtal13, Knut07}, bone density~\citep{ChilEtal12,
  BassRams94, LaynNels99}, cardiovascular
performance~\citep{MaioEtal00, PollEtal00}, lung
capacity~\citep[][although see \cite{RomaEtal16}]{LazoEtal16},
endurance~\citep{WilmKnut03}, and more.  Exercise can also
improve mental health~\citep{Ragl90, MikkEtal17, TaylEtal85,
  DeslEtal09, Call04, PaluSchw00, BassSuzu17} and cognitive performance~\citep{ChanEtal12b,
  BrisEtal02, EtniEtal06, BassSuzu17}.

@@

In the present study, we hypothesize that (a) different exercise regimens will
have different, quantifiable impacts on cognitive performance and
mental health, and that (b) these impacts will be consistant across
individuals.  To this end, we collected a year of fitness tracking
data from each of 113 participants.  We then asked each participant to
fill out a brief survey in which they self-evaluated and self-reported
several aspects
of their mental health.  Finally, we ran each participant through a
battery of memory tasks, which we used to evaluate their memory
performance along several dimensions.  We searched the data for
potential associations between memory, mental health, and exercise.

@@

    We ran an online experiment using the Amazon Mechanical Turk (MTurk)
    platform~\citep{GureEtal16}.  We collected data about each participant's fitness and
    exercise habits, a variety of self-reported measures concerning their
    mental health, and about their performance on a battery of memory
    tasks.

\subsection*{Experiment}
\subsubsection*{Participants}
We recruited experimental participants by posting our experiment as a
Human Intelligence Task (HIT) on the MTurk platform.
We limited participation to MTurk Workers who had been
assigned a ``master worker'' designation on the platform, given to workers
who score highly across several metrics on a large number of HITs,
according to a proprietary algorithm managed by Amazon.  One criteria
embedded into the algorithm is a requirement that master
workers must maintain a HIT acceptance rate of at least 95\%.  We further
limited our participant pool to participants who self-reported that
they were fluent in English and regularly used a Fitbit fitness
tracker device.  A total of 160 workers accepted our HIT in order to
participate in our experiment.  Of these, we excluded all participants
who failed to log into their Fitbit account (giving us access to their
anonymized fitness tracking data), encountered technical issues (e.g.,
by accessing the HIT using an incompatible browser, device, or
operating system), or who ended their participation prematurely,
before completing the full study.  In all, 113 participants
contributed usable data to the study.

@@

\paragraph{Reported medical conditions and medications.}
Participants reported having and/or taking medications pertaining to the following medical conditions: anxiety or
depression (4 participants), recent head injury (2 participants), high
blood pressure (1 participant), bipolar disorder (1 participant),
hypothyroidism (1 participant), and other unspecified conditions or medications (1
participant).  Participants reported their current and typical stress
levels on a Likert scale as very relaxed (-2), a little relaxed (-1),
neutral (0), a little stressed (1), or very stressed (2).  The
``current'' stress level reflected participants' stress at the time
they participated in the experiment.
Their responses
ranged from -2 to 2 (current stress: 25\textsuperscript{th} percentile: -2;
50\textsuperscript{th} percentile: -1; 75\textsuperscript{th}
percentile: 1; typical stress: 25\textsuperscript{th} percentile: 0;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 1).  Participants also reported their current level of
alertness on a Likert scale as very sluggish (-2), a little sluggish
(-1), neutral (0), a little alert (1), or very alert (2).  Their
responses ranged from -2 to 2 (25\textsuperscript{th} percentile: 0;
50\textsuperscript{th} percentile: 1; 75\textsuperscript{th}
percentile: 2).  Nearly all (111 out of 113) participants reported
that they had normal color vision, and 15 participants reported
uncorrected visual impairments (including dyslexia and uncorrected
near- or far-sightedness).

@@

\subsubsection*{Tasks}
Upon accepting the HIT posted on MTurk, each worker was
directed to read and fill out a screening and consent form, and to
share access to their anonymized Fitbit data via their Fitbit account.
After consenting to participate in our study and successfully sharing their Fitbit
data, participants filled out a survey and then engaged in a series of
memory tasks (Fig.~\ref{fig:tasks}).  All stimuli and code for running
the full MTurk experiment may be found
\href{https://github.com/ContextLab/brainfit-task}{\underline{here}}.

@@

The word lists participants studied were drawn from the categorized
lists reported by \cite{ZimaEtal18}.  Each participant was assigned
four unique randomly chosen lists (in a randomized order), selected
from a full set of 16 lists.  Each chosen list was then randomly
shuffled before presenting the words to the participants.
Participants also performed a final delayed memory test where they
were given 180~s to type out any words they remembered from
\textit{any} of the 4 lists they had studied.

@@

\paragraph*{Naturalistic recall (Fig.~\ref{fig:tasks}b).}
Participants watched a 2.5~minute video clip entitled ``The Temple of
Knowledge.''  The video comprises an animated story told to StoryCorps
by Ronald Clark, who was interviewed by his daughter, Jamilah Clark.
The narrator (Ronald) discusses growing up living in an apartment over
the Washington Heights branch of the New York Public Library, where his
father worked as a custodian during the 1940s.

After watching the video clip, participants were asked to type out
anything they remembered about what happened in the video.  They typed
their responses into a text box, one sentence at a time.  When the
participant pressed the return key or typed any final punctuation mark
(``.'', ``!'', or ``?'') the text currently entered into the box was
``submitted'' and added to their transcript, and the text box was
cleared to prevent further editing of any already-submitted text.
This was intended to prevent participants from retroactively editing
their previous responses.  Participants were given up to 10~minutes to
enter their responses.  After 4~minutes, participants were given the
option of ending the response period early, e.g., if they felt they
had finished entering all of the information they remembered.  Each
participant's transcript was constructed from their submitted
responses by combining the sentences into a single document and
removing extraneous whitespace characters.  Following this 4--10
minute free response period, participants were given a series of 10
multiple choice questions about the conceptual content of the story.
All participants received the same questions, in the same order.
Participants also performed a final delayed memory test, where they
carried out the free response recall task a second time, near the end
of the testing session.  This resulted in a second transcript, for
each participant.

\paragraph*{Foreign language flashcards (Fig.~\ref{fig:tasks}c).}
Participants studied a series of 10 English-Gaelic word pairs in a
randomized order.  We selected the Gaelic language both for its
relatively small number of native speakers and for its dissimilarity
to other commonly spoken languages amongst MTurk workers.
We verified (via self report) that all of our participants were fluent
in English and that they were neither fluent nor familiar with Gaelic.

@@

The fitness tracking data associated with participants' Fitbit
accounts varied in scope and duration according to which device the
participant owned (Fig.~\devices), how often the participant wore
(and/or synced) their tracking device, and how long they had owned
their device.  For example, while all participants' devices supported
basic activity metrics such as daily step counts, only a subset of the
devices with heart rate monitoring capabilities provided information
about workout intensity, resting heart rate, and other related
measures.  Across all devices, we collected the following information:
heart rate data, sleep tracking data, logged bodyweight measurements,
logged nutrition measurements, Fitbit account and device settings, and
activity metrics.

@@

\paragraph{Nutrition.} If available, we extracted any
nutrition-related information affiliated with participants' Fitbit
accounts within 1 year prior to enrolling in our study. Depending on
their specific account settings and usage behaviors, this included a
log of the specific foods they had eaten (and logged) over the past
year, and the amount of water consumed (and logged) each day.

@@

\subsubsection*{Comparing recent versus baseline measurements.}
We were interested in separating out potential associations between
\textit{absolute} fitness metrics and \textit{relative} metrics.  To
this end, in addition to assessing potential raw (absolute) fitness
metrics, we also defined a simple measure of recent changes in those
metrics, relative to a baseline:
\[
  \Delta_{R, B} m = \frac{B \sum_{i = 1}^R
  m(i)}{R \sum_{i=R + 1}^{R+B}m(i)},
\]
where $m(i)$ is the value of metric $m$ from $i - 1$ days prior to
testing (e.g., $m(1)$ represents the value of $m$ on the day the
participant accepted the HIT, and $m(10)$ represents the value of $m$
9 days prior to accepting the HIT.  We set $R = 7$ and $B = 30$.  In
other words, to estimate recent changes in any metric $m$, we divided
the average value of $m$ taken over the prior week by the average
value of $m$ taken over the 30 days before that.

\subsubsection*{Exploratory correlation analyses}
We used a bootstrap procedure to identify reliable correlations
between different memory-related, fitness-related, and
demographic-related variables.  For each of $N = 10,000$ iterations, we
selected (with replacement) a sample of 113 participants to include.
This yielded, for each iteration, a sampled ``data matrix'' with one
row per sampled participant and one column for each measured variable.
When participants were sampled multiple times in a given iteration, as
was often the case, this matrix contained duplicate rows.  Next, we computed the Pearson's correlation
between each pair of columns.  This yielded, for each pair of columns,
a distribution of $N$ bootstrapped correlation coefficients.  If
$97.5\%$ or fewer of the coefficients for a given pair of columns had the
same sign, we excluded the pair from further analysis and considered
the expected correlation between those columns to be undefined.  If
$> 97.5\%$ of the coefficients for a given pair of columns had the
same sign (corresponding to a bootstrap-estimated two-tailed $p$
threshold of 0.05), we computed the expected correlation coefficient as:
\[
  \mathbb{E}_{i, j}\left[ r\right] = \tanh\left(\frac{1}{N} \sum_{n=1}^N
  \tanh^{-1}(\mathrm{corr}\left(m(i)_n, m(j)_n\right))\right),
\]
where $m(x)_n$ represents column $x$ of the bootstrapped data matrix
for iteration $n$, $\tanh$ is the hyperbolic tangent, and $\tanh^{-1}$
is the inverse hyperbolic tangent.  We estimated the corresponding $p$-values
for these correlations as one minus the proportion of bootstrapped
correlations with the same sign, multiplied by two.

\subsubsection*{Reverse correlation analyses}
We sought to characterize potential associations between the \textit{dynamics}
of participants' fitness-related activities leading up to the time
they participated in a memory task and their performance on the given
task.  For each fitness-related variable, we constructed a timeseries
matrix whose rows corresponded to timepoints (sampled once per day)
leading up to the day the participant accepted the HIT for our study,
and whose columns corresponded to different participants.  These
matrices often contained missing entries, since different
participants' Fitbit devices tracked fitness-related activities
differently.  For example, participants whose Fitbit devices lacked
heart rate sensors would have missing entries for any heart
rate-related variables.  Or, if a given participant neglected to wear
their fitness tracker on a particular day, the column corresponding to
that participant would have missing entries for that day.  To create
stable estimates, we smoothed the timeseries of each fitness measure
using a sliding window of 1 week.  In other words, for each fitness
measure, we replaced the ``observed
value'' for each day with the average values of that measure (when
available) over the 7 day interval ending on the given day.

@@

Given a fitness timeseries matrix, $F$, we computed the weighted
average and weighted standard error of the mean of each row of $F$,
where the weights were given by a particular memory-related variable
(row of $M$).  For example, if $F$ contained participants' daily step
counts, we could use any row of $M$ to compute a weighted average
across any participants who contributed step count data on each day.
Choosing a row of $M$ that corresponded to participants' performance
on the naturalistic recall task would mean that participants who
performed better on the naturalistic recall task would contribute more
to the weighted average timeseries of daily step counts.
Specifically, for each row, $t$, of $F$, we computed the weighted
average (across the $S$ participants) as:
\[
\bar{f}(t) = \sum_{s=1}^S \dot{m}(s) F(t, s),
\]
where $\dot{m}$ denotes the normalized min-max scaling of $m$ (the row
of $M$ corresponding to the chosen memory-related variable):
\[
  \dot{m} = \frac{m}{\sum_{s=1}^S \hat{m}(s)},
\]
where
\[
  \hat{m} = (1 - \epsilon)\frac{m - \min(m)}{\max(m) - \min(m)} + \epsilon.
\]
Here, $\epsilon$ provides a lower bound on the influence of the
lowest-weighted participant's data.  We defined $\epsilon = 0.001$,
ensuring that the lowest-weighted participant had relatively low (but
non-zero) influence.  We computed the weighted standard error of the
mean as:
\[
\mathrm{SEM}_m\left(f(t)\right) = \frac{\left| \sum_{s=1}^S \left( F(t, s) -
    \bar{f}(t)\right) \right|}{\sqrt{S}}.
\]
When a given row of $F$ was missing data from one or more
participants, those participants were excluded from the weighted
average for the corresponding timepoint and the weights (across all
remaining participants) were re-normalized to sum to 1.  The above
procedure yielded, for each memory variable, a timeseries of weighted
average (and weighted standard error of the mean) fitness tracking
values leading up to the day of the experiment.

\section*{Results}
Before testing our main hypotheses, we examined the behavioral data
from each of four memory tasks: a random word list learning ``free
recall'' task (Fig.~\ref{fig:tasks}); a naturalistic recall task whereby participants watched
a short video and then recounted the narrative; a foreign language
``flashcards'' task; and a spatial learning task.  Each of the first
three tasks (free recall, naturalistic recall, and the flashcards
task) included both an immediate (short term) memory test and a
delayed (long term) memory test.  The spatial learning task included
only an immediate test. Participants in all four tasks exhibited
general trends and tendencies that have been previously reported in
prior work.  We were also interested in characterizing the variability
in task performance across participants.  For example, if all
participants exhibited near-identical behaviors or performance on a
given task, we would be unable to identify how memory performance on
that task varied with mental health or exercise.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/behavior_overview_immediate}
\caption{\textbf{Immediate memory tests.}  \textbf{a. Free recall.}
  Left: probability of recalling each word first as a function of its
  presentation position.  Middle: probability of transitioning between
successively recalling the word presented at position $i$, followed by
word presented at position $i + \mathrm{Lag}$.  Right: probability of
recalling each word as a function of its presentation position.  See
Figure~\frDetail~for additional details.
\textbf{b. Naturalistic recall.}  Top: 2D embedding of a 2.5~min video
clip; each dot reflects a narrative event (red denotes early events
and blue denotes later events).  Bottom: 2D embedding of the averaged
transcripts of participants' recountings of the narrative (dots: same
format as top panel).  The arrows denote the average trajectory
directions through the corresponding region of text embedding space,
for any participants whose recountings passed through that region.
Blue arrows denote statistically reliable agreement across
participants ($p < 0.05$, corrected).  See Figure~\natDetail~for
additional details.  \textbf{c. Foreign language
  flashcards.} Each bar denotes the average proportion of correctly
recalled Gaelic-English word pairs from early (first 3), late (last
3), or all (i.e., all 10) study positions.  See
Figure~\vocabDetail~for additional details.  \textbf{d. Spatial
  learning.}  Average estimation error in shape locations as a
function of the number of shapes.  See Figure~\spatialDetail~for
additional details.  All panels: error bars and error
ribbons denote bootstrap-estimated 95\% confidence intervals.  Shading
(saturation) denotes results for different subsets of participants
assigned based on their task performance (Figs.~\frDetail,
\natDetail, \vocabDetail, and \spatialDetail~provide information about
which performance metrics and values the shading reflects; in general
more saturated colors denote participants who performed better on the
given task.)  In Panel d, participants are grouped in two ways; in the
left panel, participants are grouped according to the $y$-intercepts of regression lines (estimation error as a
function of the number of shapes); in the right panel, participants
are grouped according to the slopes of the same regression lines.}
\label{fig:immediate_behavior}
\end{figure}

@@

In addition to observing substantial across-participant variability in
memory performance, we also observed substantial variability in
participants' fitness and activity metrics
(Fig.~\ref{fig:fitness_summary}).  We examined recent measurements,
averaged over the week prior to testing
(Fig.~\ref{fig:fitness_summary}a), baselined measurements (average
over the prior week, divided by the average over the preceding 30 days;
Fig.~\ref{fig:fitness_summary}b), along with more gradually varying
measures that tended to remain relatively static over timescales of
weeks to months (Fig.~\ref{fig:fitness_summary}c).
Figure~\fitDists~displays across-participant
distributions for a broad selection of these measures, and Figures~\fitDistgridImmediate,
\fitScatterImmediate, \fitDistgridDelayed, and \fitScatterDelayed~show
different participants' fitness metrics, broken down by their
performance on different memory tasks.

@@

Although the present study cannot reveal causal effects, a large prior
literature provides some insight into potential causal effects by examining the neural and cognitive effects of a variety
of exercise interventions~\citep{ChanEtal15, VidoEtal15, KamiEtal07,
  ImboEtal19, SuwaEtal17, SinhEtal21, Tomp03}.  A limitation of that
prior work is that most of these studies examine how relatively
short-term changes in exercise (e.g., on timescales of hours to days
or, rarely, weeks to months) affect a cognitive performance on single
task or aspect of mental health.  The present study examines
longer-term exercise (over a full year), and relates long-term
exercise history to performance on a variety of tasks and to a variety
of mental health dimensions.
